Modelling palaeoecological time series using generalized additive models

### 2018-08-15 manuscript.Rmd
insert flow diagram; bump date

---
title: "Modelling palaeoecological time series using generalized additive models"
author: Gavin L. Simpson
affiliation: "Institute of Environmental Change and Society, \\ University of Regina, Regina, Saskatchewan, Canada, S4S 0A2"
mail: "gavin.simpson@uregina.ca"
date: May 11, 2018
keywords: time series; generalized additive model; simultaneous interval; spline; environmental change 
bibliography: references.bib
csl: frontiers-in-ecology-and-evolution.csl
geometry: "left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm"
fontsize: 12pt
mainfont: TeX Gyre Pagella
output:
  pdf_document:
    number_sections: true
    fig_width: 8
    fig_height: 5
    fig_crop: false
    keep_tex: true
    latex_engine: xelatex
    template: my.latex
    md_extensions: +header_attributes+superscript+subscript
    includes:
      in_header: header.tex
  md_document:
    variant: markdown_github
---

@@

In the subsequent sections, I present an introduction to GAMs and discuss the issue of uncertainty in model-estimated trends, the topic of posterior simulation from a regression model and how to identify periods of significant environmental change using the first derivative of the estimated trend. Two non-standard types of spline --- adaptive smoothers and Gaussian process splines --- that are especially applicable to GAMs in the palaeoenvironmental setting are subsequently described, followed by an assessment of the the impact of age-model uncertainty on trend estimation via GAMs. Finally, I briefly discuss the application of GAM trend analysis to multivariate species abundance and compositional data.

>>>>

---
title: "Modelling palaeoecological time series using generalized additive models"
author: Gavin L. Simpson
affiliation: "Institute of Environmental Change and Society, \\ University of Regina, Regina, Saskatchewan, Canada, S4S 0A2"
mail: "gavin.simpson@uregina.ca"
date: August 15, 2018
keywords: time series; generalized additive model; simultaneous interval; spline; environmental change 
bibliography: references.bib
csl: frontiers-in-ecology-and-evolution.csl
geometry: "left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm"
fontsize: 12pt
mainfont: TeX Gyre Pagella
output:
  pdf_document:
    number_sections: true
    fig_width: 8
    fig_height: 5
    fig_crop: false
    keep_tex: true
    latex_engine: xelatex
    template: my.latex
    md_extensions: +header_attributes+superscript+subscript
    includes:
      in_header: header.tex
  md_document:
    variant: markdown_github
---

@@

In the subsequent sections, I present an introduction to GAMs and discuss the issue of uncertainty in model-estimated trends, the topic of posterior simulation from a regression model and how to identify periods of significant environmental change using the first derivative of the estimated trend. The main steps in the analysis of palaeoenvironmental time series using GAMs are illustrated in Figure \ref{fig:flowchart}. Two non-standard types of spline --- adaptive smoothers and Gaussian process splines --- that are especially applicable to GAMs in the palaeoenvironmental setting are subsequently described, followed by an assessment of the the impact of age-model uncertainty on trend estimation via GAMs. Finally, I briefly discuss the application of GAM trend analysis to multivariate species abundance and compositional data.

\begin{figure}

{\centering \includegraphics[width=0.4\linewidth]{manual-figures/flow-diagram}

}

\caption{Flowchart showing the main steps in the analysis of time series using generalized additive models.}\label{fig:flowchart}
\end{figure}
